Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-05-15 manuscript/manuscript.tex
Various fixes
1) Fix SI package warning by adding \bar explicitly
2) Clarify intro sentence on link between free energy of solvation and hydrophobicity
3) Clarify use of restraints

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{float}
\usepackage{placeins}
\DeclareSIUnit\Molar{M}
\DeclareSIUnit\calorie{cal}
\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
%\DeclareSIUnit{\Molar}{M}
\DeclareSIUnit\angstrom{\text{\AA}}
%\DeclareSIPrefix\micro{\text{\textmu}}{0}
\sisetup{
	%math-rm=\mathsf,
	text-rm=\sffamily
}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{figures/carbazole-figure.png}
	\caption{Schematic showing regions of high water density around a carbazole solute molecule. Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 6-dimensional orientational and solvational solvent distribution $\rho$ around the solute molecule. Note that while this depiction focuses on regions of high solvent density, regions with lower density can also contain meaningful contributions to the overall energy and entropy.}
	\label{fig:carbazole}
\end{figure}
\section{Introduction}
Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.

@@

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is not well-defined for moving solute molecules.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. Note that this is a rather strong restraint. While lower values are possible, make sure that the solute molecule is kept rigid during the simulation. As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{float}
\usepackage{placeins}
\DeclareSIUnit{\bar}{bar}
\DeclareSIUnit\Molar{M}
\DeclareSIUnit\calorie{cal}
\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
%\DeclareSIUnit{\Molar}{M}
\DeclareSIUnit\angstrom{\text{\AA}}
%\DeclareSIPrefix\micro{\text{\textmu}}{0}
\sisetup{
	%math-rm=\mathsf,
	text-rm=\sffamily
}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{figures/carbazole-figure.png}
	\caption{Schematic showing regions of high water density around a carbazole solute molecule. Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 6-dimensional orientational and solvational solvent distribution $\rho$ around the solute molecule. Note that while this depiction focuses on regions of high solvent density, regions with lower density can also contain meaningful contributions to the overall energy and entropy.}
	\label{fig:carbazole}
\end{figure}
\section{Introduction}
Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
It is directly linked to the solubility of solutes and drives many real-world effects. 
For example, solvation thermodynamics govern aqueous solubility and hydrophobic effects, which in turn influence phenomena like partitioning between environments, ligand binding, and protein folding.

@@

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is defined for rigid solute molecules.
In practice, this means that the solute molecule should not move significantly during the MD simulation.
This is tipically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure). 
In AMBER,  the harmonic restraint potential takes the form $U = k_r * (r - r_0)^2$ for each cartesian coordinate of each restrained atom, where $r_0$ is its reference position. 
The force constant $k_r$ should be chosen strong enough to prevent significant deviation.
Here, we apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
Note that this is a rather strong restraint. 
While weaker restraints are possible, make sure that the solute molecule is kept rigid during the simulation. 
As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
Similarly, the solute could also be kept rigid using positional constraints, RMSD restraints, or other methods.
However, depending on the method, additional post-processing may be required to keep the solute at the same position in the grid.
To make sure that the solute is not moving significantly, we recommend to check the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}
